### 2025-05-26 paper/full.tex
Update brandl results.

\setcounter{proposition}{0}
\setcounter{assumption}{0}
\setcounter{lemma}{0}
\setcounter{theorem}{0}

>>>>

\documentclass[a4paper,12pt]{report}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

%% Sets page size and margins according to Bonn University guidelines
\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=2cm]{geometry}

%% Font settings - Times New Roman or similar (as per guidelines)
\usepackage{times} % Times font
\usepackage{mathptmx} % Times font with math support

%% Section formatting
\usepackage{titlesec}

% Traditional academic chapter title style
\titleformat{\chapter}[display]
  {\normalfont\Large\bfseries}
  {}
  {0pt}
  {\thispagestyle{empty}%
   \vspace*{-20pt}%
   \begin{flushright}
     % Elegant, academic chapter number
     \begin{tikzpicture}[remember picture, overlay]
       \node[text=black!12, font=\fontsize{80}{80}\selectfont] {\thechapter};
     \end{tikzpicture}
   \end{flushright}
   \vspace{-60pt}
   % Simple, elegant single line
   \rule{\textwidth}{0.6pt}\\
   \vspace*{15pt}
   \raggedright\LARGE\scshape}
  [\vspace{-0.8cm}]
\titlespacing*{\chapter}{0pt}{0pt}{40pt}

% Elegant section formatting
\titleformat{\section}
  {\normalfont\Large\bfseries\scshape}
  {}
  {0pt}
  {\thesection\quad}
\titlespacing*{\section}{0pt}{3.5ex plus 1ex minus .2ex}{1.5ex plus .2ex}

% Make subsections italic and elegant
\titleformat{\subsection}
  {\normalfont\large\bfseries\itshape}
  {\thesubsection}
  {1em}
  {}
\titlespacing*{\subsection}{0pt}{3ex plus 1ex minus .2ex}{1.5ex plus .2ex}

%% Line spacing 1.5 as required
\usepackage{setspace}
\onehalfspacing

%% Useful packages
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}         % For bold symbols
\usepackage{graphicx}
\usepackage{float}      % For [H] figure placement
\usepackage{booktabs}
\usepackage{tabu}
\usepackage{enumitem}
\usepackage[table,dvipsnames]{xcolor}  % Added dvipsnames option for more color choices
\usepackage{array}
\usepackage{multirow}
\usepackage{subfigure}
\usepackage{amsthm}
\usepackage{bbm}        % Blackboard math symbols (for indicator function)
\usepackage[toc]{appendix}

% Choose either algorithm2e OR algorithmicx, not both
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}

% For adjusting margins of specific sections
\usepackage{changepage}

% Bibliography
\usepackage[natbibapa]{apacite}

% Todo notes
\usepackage[colorinlistoftodos]{todonotes}

% Page numbering configuration
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0pt}

% Hyperref should be last
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

% Allow floating content to be placed more liberally
\renewcommand{\floatpagefraction}{.9}
\renewcommand{\textfraction}{.1}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{.9}

% Appendix formatting
\makeatletter
\let\old@chapter\@chapter
\def\@chapter[#1]#2{\old@chapter[{#1}]{#2}}
\makeatother

% Math operators
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

% Custom theorem environments
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
\newtheorem{assumption}{Assumption}
\newtheorem{remark}{Remark}

% Define indicator function notation
\newcommand{\ind}[1]{\mathbbm{1}_{#1}}

\begin{document}

% Cover sheet as per Bonn University guidelines
\begin{titlepage}
    \begin{center}
        \vspace*{2cm}

        {\Large \textbf{Social Learning via Multi Agent Reinforcement Learning}}\\[6cm]

        Master Thesis Presented to the\\
        Department of Economics at the\\
        Rheinische Friedrich-Wilhelms-Universität Bonn\\[1cm]

        In Partial Fulfillment of the Requirements for the Degree of\\
        Master of Science (M.Sc.)\\[4cm]

        Supervisor: Prof. Dr. Florian Brandl\\[2cm]

        Submitted in June, 2025 by:\\
        Ege Can Doğaroğlu\\
        Matriculation Number: 3464688

    \end{center}
\end{titlepage}

% Set page numbering to 'gobble' (no numbers) for front matter
\pagenumbering{gobble}

% Abstract
\begin{abstract}
    \begin{adjustwidth}{2cm}{2cm}
        This thesis bridges economic social learning theory and multi-agent reinforcement learning by introducing Partially Observable Active Markov Games (POAMGs)—a novel framework for modeling sophisticated learning dynamics in environments with partial observability. Traditional economic models provide valuable theoretical insights but often rely on simplified assumptions about agents' behavior or the learning environment, while reinforcement learning approaches offer computational tools that frequently lack theoretical foundations for strategic adaptation. Our framework addresses three fundamental challenges: modeling strategic adaptation in non-stationary environments, incorporating partial observability as an intrinsic feature, and capturing long-term strategic considerations in learning contexts. We develop POLARIS (Partially Observable Learning with Active Reinforcement In Social environments), a practical algorithm implementing our theoretical framework, and apply it to two canonical social learning scenarios: strategic experimentation with observable rewards and learning without experimentation through action observation. Our results validate theoretical predictions about ... while demonstrating how reinforcement learning agents can discover sophisticated strategies for information revelation and strategic influence, providing both theoretical insights and computational tools for understanding complex social learning dynamics.
    \end{adjustwidth}
\end{abstract}

% Table of contents
\tableofcontents
\newpage

% Optional - include if you have abbreviations, tables, or figures to list
%\listoffigures
%\listoftables
%\newpage

% Page numbering starts here with Arabic numerals
\chapter{Introduction}
\pagenumbering{arabic}
\setcounter{page}{1}
The study of how individuals and groups learn from each other's actions and experiences has been a central focus in economic theory for decades. From the seminal contributions on information cascades by \citet{banerjee1992simple} and \citet{bikhchandani1992theory} to more recent explorations of learning in networks \citep{acemoglu2011bayesian, golub2010naive}, economic research has sought to understand how social interactions shape beliefs, decisions, and collective outcomes. Traditional economic models have provided valuable insights into phenomena such as herding behavior, where individuals rationally ignore their private information to follow the observed actions of predecessors. However, these models often rely on simplified assumptions about agents' behavior and reasoning processes, frequently restricting analysis to one-shot sequential decisions or steady-state equilibria rather than capturing the rich dynamics of repeated interactions where agents continuously adapt their strategies based on others' evolving behaviors.

Meanwhile, the field of multi-agent reinforcement learning (MARL) has experienced remarkable progress in developing algorithms that enable autonomous agents to learn effective strategies in complex, interactive environments. Recent advances in reinforcement learning have demonstrated impressive capabilities in strategic games \citep{silver2016mastering}, cooperative problem-solving \citep{baker2020emergentreciprocityteamformation}, and competitive scenarios \citep{openai2019dota2largescale}.  MARL offers powerful tools for modeling adaptive behavior in non-stationary environments where agents must continuously revise their strategies in response to others' changing behaviors. However, much of this work has focused on engineering objectives rather than modeling realistic human learning processes, limiting its applicability to fundamental questions in economic theory. Additionally, many MARL approaches struggle with the challenge of partial observability—a defining feature of social learning contexts where agents cannot directly observe others' private information or belief states.

This thesis bridges these domains by introducing a novel framework that integrates economic social learning theory with multi-agent reinforcement learning under partial observability. Our approach models social learning as a dynamic process where agents with limited information strategically adapt their behaviors while simultaneously learning about their environment and other agents' strategies. By formalizing this process through the lens of partially observable reinforcement learning, we aim to reveal new insights into how strategic adaptation influences collective outcomes and equilibrium behavior. This interdisciplinary synthesis addresses fundamental challenges that have constrained previous work in both economics and artificial intelligence, offering a more realistic and computationally tractable approach to modeling complex social learning dynamics.

The first challenge concerns non-stationarity and strategic adaptation. Traditional economic models of social learning typically assume that agents follow fixed, myopic decision rules or Bayesian updating procedures that do not fully account for strategic adaptation over time. When agents repeatedly interact and observe each other's actions, however, they may adjust their strategies in anticipation of others' learning, creating a complex web of interdependent adaptation that fundamentally alters collective learning dynamics. This non-stationarity—where the effective environment an agent faces changes as other agents learn—represents a central challenge that requires new modeling approaches. Standard MARL algorithms often treat this non-stationarity as a technical obstacle to convergence rather than an intrinsic feature of multi-agent systems that agents should explicitly reason about and strategically exploit.

The second challenge involves partial observability, which pervades social learning contexts. In most real-world settings, agents cannot directly observe others' private information, beliefs, or decision processes. Instead, they must infer these hidden states from observable actions and outcomes. This partial observability compounds the complexity of strategic interaction, as agents must simultaneously learn about the underlying state of the world and about the belief formation processes of others. Economic models have often simplified this challenge through strong assumptions about common knowledge or by focusing on one-shot sequential decisions rather than repeated interactions. Similarly, many MARL approaches assume full observability of the state or treat partial observability as a technical challenge to be addressed through belief state tracking, without fully incorporating its strategic implications.

The third challenge concerns long-term strategic considerations in social learning. Agents often face a tension between immediate payoffs and long-term information generation. They may sometimes choose actions that appear suboptimal in the short term to influence the learning trajectories of others or to generate valuable information for future periods. For instance, in strategic experimentation contexts, agents might incur costs to explore unknown options, knowing that the resulting information will benefit both themselves and others in the future. Capturing these farsighted strategic considerations requires models that explicitly account for how current actions shape the future evolution of others' beliefs and behaviors—a capability that neither traditional economic models nor standard MARL algorithms fully provide.

To address these challenges, this thesis makes several interconnected contributions. First, we develop Partially Observable Active Markov Games (POAMGs)—a novel formalism that extends the Active Markov Game framework introduced by \citet{kim2022influencing} to partially observable settings. Unlike standard reinforcement learning frameworks that treat non-stationarity as a challenge to be mitigated, POAMGs incorporate policy evolution as an integral part of the environment dynamics, allowing agents to reason about and strategically influence this evolution process. POAMGs explicitly model how agents' policies evolve over time based on observations and interactions, while accounting for the fundamental constraints imposed by partial observability. This formalism provides a mathematically rigorous foundation for analyzing complex social learning dynamics that involve repeated interactions, adaptive strategies, and incomplete information.

Second, we provide a thorough theoretical analysis of convergence and equilibrium properties in POAMGs. We establish conditions under which the joint process of states, beliefs, and policy parameters converges to a unique stochastically stable distribution, ensuring that our models have well-defined limiting behavior despite the inherent non-stationarity of multi-agent learning. We derive policy gradient theorems for average and discounted reward objectives, providing a solid foundation for algorithm development. We also extend our framework to continuous-time dynamics through stochastic differential equations, enabling a wider range of applications. Our analysis illuminates the relationship between traditional game-theoretic equilibrium concepts and the more general notion of active equilibrium that emerges in our framework, offering new insights into equilibrium selection and stability in social learning contexts.

Third, we introduce POLARIS (Partially Observable Learning with Active Reinforcement In Social environments), a practical algorithm for learning in POAMGs. POLARIS extends previous algorithms to partially observable settings through an integrated architecture with three core components: a belief processing module that tracks agents' information states using Transformer models; an inference learning module that predicts the policy evolution of other agents through variational methods; and a reinforcement learning module that optimizes policies based on average or discounted reward criteria, supporting both discrete and continuous action spaces. This integrated approach enables agents to learn sophisticated strategies that account for both the partial observability of the environment and the learning dynamics of other agents—capabilities that are essential for modeling realistic social learning processes.

Fourth, we apply our framework to two canonical social learning scenarios from economic theory. The first application focuses on strategic experimentation, building on the models of \citet{bolton1999strategic} and \citet{keller2020undiscounted}, where agents learn optimal actions through trial and error with observable rewards. The second application examines learning without experimentation, extending the frameworks of \citet{huang2024learning} and \citet{brandl2024}, where agents form beliefs primarily by observing others' actions rather than receiving direct feedback from the environment. Our analysis demonstrates how our approach can model phenomena such as free-riding on others' information production, strategic teaching, information cascades, and confounded learning under more realistic assumptions about agents' reasoning capabilities and strategic sophistication.

Through these applications, we illustrate how bridging economic theory and MARL can enhance our understanding of social learning processes. Our POAMG framework provides a more realistic model of how partially rational agents learn from each other in complex, partially observable environments. At the same time, it offers practical algorithms for computing equilibrium strategies that traditional economic approaches might find intractable. The POLARIS algorithm demonstrates how techniques from deep reinforcement learning and approximate inference can be adapted to tackle the specific challenges of social learning, offering a computationally feasible method for simulating and analyzing complex multi-agent learning dynamics.

The remainder of this thesis is organized as follows: Chapter 2 reviews the relevant literature on social learning in economic theory and multi-agent reinforcement learning, emphasizing the complementary strengths and limitations of these approaches and identifying opportunities for cross-fertilization. Chapter 3 introduces our theoretical framework, developing the formalism of Partially Observable Active Markov Games and deriving key results on convergence, policy gradients, and equilibrium concepts, as well as presenting the POLARIS algorithm, detailing its components, implementation considerations, and theoretical justifications. Chapter 4 applies our framework to strategic experimentation and learning without experimentation, demonstrating how POLARIS captures sophisticated social learning dynamics in these contexts. Chapter 5 concludes with a discussion of implications, limitations, and directions for future research.

By integrating perspectives from economic theory and multi-agent reinforcement learning, this thesis aims to enhance our understanding of social learning processes and provide new tools for modeling complex strategic interactions in partially observable environments. The insights gained may not only advance theoretical knowledge but also inform the design of platforms and institutions that facilitate efficient collective learning and decision-making in diverse contexts, from financial markets and organizational learning to online social networks and collaborative scientific endeavors. Moreover, by demonstrating the value of MARL techniques for addressing classical questions in economic theory, we hope to encourage further cross-disciplinary work that leverages complementary insights from these rich intellectual traditions.

In the next chapter, we provide a detailed review of the relevant literature, tracing the development of social learning models in economics and recent advances in multi-agent reinforcement learning. This review establishes the conceptual foundation for our theoretical framework and highlights the specific gaps that our approach aims to address, setting the stage for the formal development of Partially Observable Active Markov Games in Chapter 3.
\chapter{Related Work}
This section reviews key literature across economic social learning and multi-agent reinforcement learning (MARL), highlighting how our approach bridges these fields to address their respective limitations.

\section{Social Learning}
The economic study of social learning originated with \citet{banerjee1992simple} and \citet{bikhchandani1992theory}, who formalized how rational agents might ignore private information to follow predecessors' actions, leading to information cascades and potentially inefficient herding. While foundational, these models rely on one-shot sequential decisions rather than the repeated interactions that characterize many real-world learning contexts. \citet{smith2000pathological} extended this work by showing how heterogeneous preferences can lead to confounded learning, where private signals remain relevant despite observing others' actions.

Social learning in network settings expands this framework by examining how network structure influences information flow \citep{acemoglu2011bayesian, golub2010naive}. Two principal approaches have emerged: Bayesian models where agents perform rational inference \citep{rosenberg2009informational, gale2003bayesian,mossel2015strategic}, and non-Bayesian models \citep{golub2010naive, demarzo2003persuasion} like the DeGroot framework \citep{degroot1974reaching} where agents update beliefs through weighted averaging of neighbors' opinions. A critical insight from this literature is that network topology significantly affects learning outcomes. However, most network models still fall short in capturing how agents strategically adapt to others' evolving learning behaviors over time—a key element of our framework. subsectionWhile early models focused on one-shot decisions, more recent work has explored repeated interactions where agents continuously adapt strategies based on observations. These settings more closely align with our MARL approach and will be the focus of our implementation.

\subsection{Strategic Experimentation}

The strategic experimentation literature examines settings where agents balance exploiting current knowledge against generating new information through exploration \citep{bolton1999strategic, keller2005strategic}. This creates a dynamic tension between individual incentives to free-ride on others' information production and collective benefits from experimentation. Strategic experimentation represents a fundamental departure from classical social learning models by explicitly accounting for the intertemporal nature of information acquisition. Unlike cascade models where agents make one-shot decisions in sequence, agents in strategic experimentation scenarios face repeated opportunities to learn and adapt their strategies over time. This dynamic perspective connects directly to the reinforcement learning paradigm, where exploration-exploitation tradeoffs are central \citep{sutton2018reinforcement}. These tradeoffs create the necessary tension for social influence, even in the absence of informational asymmetry among agents that is often central to social learning models \citep{gale2003bayesian}.

The economic foundations of strategic experimentation were established by \citet{rothschild1974two}, who analyzed how a monopolist might experiment with different prices to learn about demand. This concept was extended to multi-agent settings by \citet{bolton1999strategic}, who developed a framework for analyzing experimentation in teams. Their seminal work revealed that when information is a public good, free-riding incentives can significantly reduce aggregate experimentation below socially optimal levels, creating a classic collective action problem.

Several extensions have explored how different information structures affect experimentation incentives. \citet{keller2005strategic} introduced exponential bandits, where lump-sum rewards arrive according to a Poisson process, demonstrating how the resolution of uncertainty affects the dynamics of experimentation. Their model showed that "encouragement effects" can arise, where agents experiment more intensively to motivate others to join the exploration effort. \citet{klein2011negatively} demonstrated how negatively correlated bandits—where success on one experiment decreases the estimated value of others—can encourage more efficient experimentation patterns.
\iffalse
    \citet{bonatti2017dynamic} further showed how competition in experimentation can sometimes overcome free-riding tendencies and increase information production through strategic preemption.\fi

\citet{keller2020undiscounted} developed a particularly relevant model using average reward criteria rather than discounted objectives. Under this framework, the value of information doesn't decay over time, incentivizing different patterns of exploration. This approach aligns with our POAMG framework's emphasis on long-term strategic adaptation in multi-agent systems. \citet{heidhues2015strategic} further demonstrated how private observations can restore experimentation incentives that fail under public observations, providing insights into how information asymmetry affects collective learning dynamics.

The strategic teaching phenomenon, where agents take seemingly suboptimal actions to influence others' beliefs, emerges naturally in these contexts. \citet{yamamoto2019stochastic} demonstrated how sophisticated agents might deliberately punish or reward others via continuation payoffs due to invariance property of the payoff sets. Similarly, \citet{halac2017designing} showed how optimal incentive structures might intentionally incentivize agents to experiment, highlighting the importance of mechanism design in collective learning environments.

The literature on information design \citep{kamenica2011bayesian, bergemann2019information} provides complementary insights by examining how information revelation mechanisms affect experimentation decisions. \citet{che2018optimal} showed how over-revealing information can sometimes incentivize more experimentation than full transparency. These insights connect directly to the strategic influence aspects of our POAMG framework, which explicitly models how agents reason about and deliberately influence others' learning trajectories.

\subsubsection{Formal Model}
More formally, we can describe the strategic experimentation literature through the model developed by \citet{keller2020undiscounted}. In this framework, $n$ agents face a two-armed bandit problem where they continuously allocate resources between a safe arm with known deterministic payoff $r_\textit{safe} > 0$ and a risky arm whose expected payoff depends on an unknown state $\omega \in \{0,1,\ldots,m\}$ with $m \geq 1$. The state is drawn at the beginning according to a publicly known prior distribution with full support. The risky arm generates payoffs according to a Lévy process:
\begin{equation*}
    X^i_t = \alpha_{\omega} t + \sigma Z^i_t + Y^i_t
\end{equation*}
where $Z^i$ is a standard Wiener process, $Y^i$ is a compound Poisson process with Lévy measure $\nu_{\omega}$, and $\alpha_{\omega}$ is the drift rate in state $\omega$. The expected payoff per unit of time is $r_{\omega} = \alpha_{\omega} + \lambda_{\omega} h_{\omega}$, where $\lambda_{\omega}$ is the expected number of jumps per unit of time and $h_{\omega}$ is the expected jump size.

At each moment, agent $i$ allocates fraction $a^i_{t} \in [0,1]$ to the risky arm, yielding instantaneous expected payoff:
\begin{equation*}
    (1-a^i_{t})r_\textit{safe} + a^i_{t}r_{\omega}
\end{equation*}

Each agent observes their own payoff process, the payoff processes of all other agents, and potentially a background information process $B_t$, which provides free information about the state. This background process follows the same structure as the payoff processes:
\begin{equation*}
    B_t = \beta_{\omega} t + \sigma_B Z^B_t + Y^B_t
\end{equation*}
whose informativeness is given exogenously as $k_0$ which increases linearly in time. Under the strong long-run average criterion, agents maximize:
\begin{equation}
    \lim_{T \to \infty} \mathbb{E}\left[\frac{1}{T}\int_0^{T}\{(1-a^i_{t})r_\textit{safe} + a^i_{t}r_{\omega}\}dt\right]
\end{equation}

The unique symmetric Markov perfect equilibrium strategy is characterized as:

\begin{equation}
    a^*(b) =
    \begin{cases}
        0                    & \text{if } I(b) \leq k_0,            \\
        \frac{I(b)-k_0}{n-1} & \text{if } k_0 < I(b) < k_0 + n - 1, \\
        1                    & \text{if } I(b) \geq k_0 + n - 1.
    \end{cases}
\end{equation}

where $I(b)$ corresponds to the incentive to experiment, defined as:
\begin{equation*}
    I(b) = \frac{f(b) - r_\textit{safe}}{r_\textit{safe} - m(b)}
\end{equation*}
when $m(b) < s$, and $I(b) = \infty$ otherwise, where $f(b)$ is the expected flow payoff under full information and $m(b)$ is the expected flow payoff from the risky arm.

This formulation directly captures the tension between exploitation (choosing the currently optimal action) and exploration (generating information for future decisions), as well as the free-rider problem that arises when information is a public good. While strategic experimentation models capture important elements of learning dynamics, they typically assume perfectly rational agents and often rely on standardized information structures.

\subsection{Learning Without Experimentation}

A complementary strand of research examines settings where agents learn without directly observing rewards. \citet{huang2024learning} studied how long-lived agents learn in networks through repeated interactions, revealing a fundamental inefficiency: regardless of network size, learning speed remains bounded by a constant dependent only on private signal distributions. \citet{brandl2024} extended these results by showing that this limitation doesn't apply uniformly to all agents, constructing scenarios where some agents learn faster at others' expense.

Unlike strategic experimentation models where agents receive direct payoff feedback, learning without experimentation captures scenarios where agents must form beliefs based primarily on others' observed actions. This distinction is crucial for modeling many real-world social learning contexts, from financial markets \citep{avery1998multidimensional} to technology adoption \citep{ellison1993rules}, where payoffs are delayed, noisy, or unobservable.

The theoretical foundations for this approach draw from both Bayesian and non-Bayesian learning traditions. \citet{gale2003bayesian} and \citet{smith2000pathological} developed early models showing how rational agents might become trapped in information cascades when learning from others' actions. \citet{acemoglu2011bayesian} extended this analysis to network settings, demonstrating how network topology influences the aggregation of dispersed information.

A substantial literature has explored learning rates in networked environments. \cite{bala1998learning} show neighborhood structure influences optimal action adoption, while \cite{golub2012homophily} demonstrate that homophily slows consensus convergence without being affected by network density. \cite{Harel2014TheSO} quantify information loss when observing others' discrete actions, finding only a fraction of private information transmits, approaching zero in large societies due to "groupthink." \cite{Jadbabaie2013InformationHA} characterize learning rates through agents' signal structures and eigenvector centralities, showing optimal information allocation depends on its distribution—better information should be placed at central nodes when information structures are comparable, but at peripheral nodes when agents possess unique critical information.

The mechanisms behind these learning barriers stem from information loss in the action quantization process. When continuous beliefs are compressed into discrete actions, information is inherently lost \citep{smith2000pathological}. \citet{guarino2013social} characterized this as 'coarse inference' where agents make inferences based only on the aggregate distribution of actions across states rather than on the fine details of how actions depend on specific histories, leading to a loss of information.

\iffalse
    Recent work has examined how behavioral factors and network structure interact to shape learning outcomes. \citet{molavi2018theory} developed models of "rational inattention" where agents optimally allocate limited cognitive resources to process social information. \citet{dasaratha2020learning} showed how certain network structures can amplify or mitigate the effects of confirmation bias and other cognitive distortions. \citet{jadbabaie2012non} proposed hybrid learning models that combine Bayesian updating with naive opinion averaging, better matching empirical patterns of social influence.
\fi

Strategic considerations emerge naturally in these settings as agents realize their actions influence the future learning of others. \citet{bhattacharya2013strategic} demonstrated how forward-looking agents might distort their actions to manipulate the information revealed to others. \citet{ARIELI2019185} showed how agents with private information might strategically time their actions to maximize influence on others' beliefs. These strategic dynamics align closely with the active influence mechanisms in our POAMG framework.

\subsubsection{Formal Model}
Similar to before, we characterize the learning dynamics in settings without experimentation through the model introduced by \citet{brandl2024}. In this framework, a set of $N= \{1, \ldots, n\}$ agents interact over discrete time periods $t \in \{1, 2, \ldots\}$ in a fixed social network. The state of the world $\omega$ is drawn from a finite set $S$ according to a prior distribution with full support and remains fixed throughout.

At each period $t$, each agent $i$ receives a private signal $o^i_t$ from a set $O$, drawn according to a state-dependent public known distribution. Signals are independent across agents and time periods. Each agent then chooses an action $a^i_t$ from a set $A$, observing the actions taken by neighbors $N^i \subset N$ in previous periods.

All agents share the same utility function $u: S \times A \rightarrow \mathbb{R}$, which depends on the state and their own action. For each state $s$, there is a unique optimal action $a_{s} = \arg\max_{a \in A} u(s, a)$, and no action is optimal in two different states. Crucially, agents do not observe their realized utilities, eliminating experimentation motives. Agent $i$'s information set at period $\tau$ consists of:
\begin{equation*}
    I^i_{\leq \tau} = (o^i_1, \ldots, o^i_{\tau}; (a^j_t)_{j \in N^i, t < \tau})
\end{equation*}

A (pure) strategy for agent $i$ is a function $\sigma^i$ that maps information sets to actions:
\begin{equation*}
    \sigma^i: \cup_{t=1}^{\infty} (O^t \times A^{N^i \times (t-1)}) \rightarrow A
\end{equation*}

For any given strategy profile $\boldsymbol{\sigma} = (\sigma^1, \ldots, \sigma^n)$, the learning rate of agent $i$ is defined as:
\begin{equation*}
    r^i(\boldsymbol{\sigma}) = \lim\inf_{t\rightarrow\infty} -\frac{1}{t}\log \mathbb{P}(a^i_t \neq a_{\omega} | \boldsymbol{\sigma})
\end{equation*}

The main theoretical results establish both limitations and opportunities in social learning networks. First, the social learning barrier theorem demonstrates that regardless of network structure or strategies, some agent's learning rate is bounded by a constant that depends only on the signal structure:
\begin{equation*}
    \min_{i \in n} r^i(\boldsymbol{\sigma}) \leq r_{bdd} = \min_{\theta \neq \theta'} \left\{D_{KL}(\mu_{\theta} || \mu_{\theta'}) + D_{KL}(\mu_{\theta'} || \mu_{\theta})\right\}
\end{equation*}
where $D_{KL}(\mu_{\theta} || \mu_{\theta'})$ is the Kullback-Leibler divergence between signal distributions and the summation corresponds to the Jeffreys divergence \citet{52862b84-7c34-3186-8b28-6af4806272c3}. This bound applies regardless of the network size or structure. Second, the coordination benefit theorem establishes that for large enough networks and for any $\epsilon > 0$, all agents can learn at rates above a certain bound:
\begin{equation*}
    \min_{i \in n} r^i(\boldsymbol{\sigma}) \geq r_{crd} - \epsilon
\end{equation*}
where $r_{crd} = \min_{\theta \neq \theta'} D_{KL}(\mu_{\theta} || \mu_{\theta'})$ is strictly greater than the autarky learning rate that an agent can achieve alone. This theoretical framework illuminates a fundamental challenge in social learning: information aggregation can fail due to information cascades and herding dynamics, resulting in, as pointed out by the paper, what \citet{harel2021rational} termed "rational groupthink"—where feedback loops cause agents to persist in incorrect beliefs despite continual information arrival. While this model allows agents to learn from various observation structures, it does not impose full rationality on the agents, a strand of literature that we discuss next.

\subsection{Non-Bayesian Learning}

While most economic models assume fully rational agents, partial rationality perspectives acknowledge cognitive limitations that affect learning. Models like \citet{jadbabaie2012non}'s hybrid learning rule (combining Bayesian updating with naive averaging) and level-$k$ reasoning \citep{stahl1994experimental, crawford2007level} provide middle grounds between full rationality and simple heuristics.
\iffalse
    Research on information processing constraints, such as bounded memory \citep{monte2014bounded} and rational inattention \citep{sims2003implications}, further illustrates how cognitive limitations shape learning outcomes. Behavioral biases like confirmation bias \citep{rabin1999psychology}, correlation neglect \citep{enke2019correlation}, and naive herding \citep{eyster2010naive} additionally affect how agents process social information. These behavioral factors often exacerbate herding behavior and information cascades, creating a gap between theoretical predictions based on full rationality and observed behavior. Our MARL framework provides a computational approach to modeling partial rationality that captures these behavioral patterns while maintaining analytical tractability.
\fi

Evolutionary Game Theory (EGT) also provides a powerful non-Bayesian framework for modeling multi-agent learning dynamics without assuming full rationality. Instead, EGT examines how strategy distributions evolve through selection processes based on relative performance \citep{weibull1997evolutionary}. This perspective aligns naturally with reinforcement learning's trial-and-error approach. \citet{tuyls2004evolutionary} demostrates several important connections between EGT and MARL among which, they relate multi-agent Q-learning to replicator dynamics. This mathematical equivalence provides theoretical insights into MARL convergence and equilibrium properties, helping explain empirical observations in complex social dilemmas \citep{leibo2017multi}.

Bridging theoretical frameworks and computational implementations, MARL offers a powerful methodology for operationalizing social learning models. MARL provides computational tools that can simulate the intricate dynamics of social learning environments while implementing the strategic adaptations that economic theories recognize as essential but frequently find difficult to compute in complex, realistic scenarios.

\section{Multi-Agent Reinforcement Learning}

While economic models offer valuable theoretical insights, they often struggle with computational tractability when modeling repeated strategic interactions with partially rational agents. This is where MARL provides complementary tools that address specific limitations of economic approaches. MARL frameworks enable the simulation of complex multi-agent systems where agents learn optimal policies through trial-and-error interactions with their environment and other agents. Unlike traditional economic models that often require closed-form solutions, MARL can handle high-dimensional state spaces, complex agent interactions, and non-stationary dynamics that emerge when multiple agents learn simultaneously.

The core components of MARL include state representations, action spaces, reward functions, and learning algorithms that enable agents to maximize expected cumulative rewards. These components can be tailored to model various aspects of social learning, including partial observability (through belief states), strategic adaptation (through policy gradient methods), and partial rationality (through constrained optimization). Modern MARL approaches incorporate techniques such as centralized training with decentralized execution, value decomposition, and multi-agent actor-critic methods to address coordination challenges that arise in multi-agent settings. We direct the interested reader to Appendix \ref{appendix:rl_details} for a detailed introduction of reinforcement learning techniques, including formal definitions, algorithms, and theoretical properties.

Economic models typically assume either fully rational agents (in Bayesian frameworks) or overly simplistic learning rules (in behavioral models). MARL offers a middle ground by modeling agents that learn from experience and adapt over time without requiring full rationality. \citet{ndousse2021emergent} demonstrated that even without explicit programming, reinforcement learning agents can develop sophisticated social learning capabilities that mirror human behavior. This addresses the partial rationality problem by providing computational mechanisms for flexible belief updating based on partial information, learning complex strategies through trial and error, and adapting to non-stationary environments created by other learning agents.

A key limitation of economic social learning models is their difficulty in capturing how agents strategically adapt to others' learning processes. The strategic experimentation literature acknowledges these dynamics but often lacks tractable solutions outside of simplified settings. \citet{jaques2019social} addressed this by introducing Social Influence as a mechanism in MARL, where agents receive additional reward for causally influencing others' actions. This creates a computational framework for modeling strategic teaching and information revelation—key dynamics in the economic models of \citet{bolton1999strategic} and \citet{heidhues2015strategic}.

More directly relevant to our approach, \citet{kim2022influencing} developed Active Markov Games, which explicitly model how agents reason about and influence the policy evolution of other agents. This formalism allows us to capture the strategic adaptation that economic models identify as important but struggle to compute in complex environments.

Despite these advantages, standard MARL approaches have their own limitations when applied to social learning. Most MARL algorithms assume full observability of state information, while social learning inherently involves partial observability of others' private information and beliefs. Many MARL approaches treat non-stationarity as a technical obstacle rather than a strategic feature to be exploited. Additionally, MARL often lacks the theoretical foundations that economic models provide for understanding equilibrium behavior.

Our POAMG framework extends Active Markov Games to partially observable settings specifically to address these limitations. By incorporating policy evolution as an integral part of the environment dynamics while accounting for partial observability, we provide a computational approach that preserves the strategic sophistication of economic models.

\section{Bridging Economic Theory and MARL}

While economic social learning and MARL have developed largely in parallel, their complementary strengths suggest significant potential for integration. Economic models provide rigorous theoretical foundations for understanding rational behavior, belief formation, and information aggregation in social contexts. However, these models often face computational limitations when addressing complex strategic interactions, especially when agents have heterogeneous beliefs, partial rationality, or operate in environments with partial observability.

Conversely, MARL offers computational frameworks for modeling adaptive agents in complex, high-dimensional environments. These approaches excel at simulating emergent behaviors and can operate effectively without imposing full rationality assumptions. However, MARL approaches frequently lack the theoretical grounding to interpret equilibrium properties and sometimes overlook the strategic sophistication captured in economic models.

Our research bridges these fields by developing a partially observable active Markov game (POAMG) framework that preserves the strategic considerations central to economic theory while leveraging the computational scalability of MARL. This integration addresses three key challenges:

First, we explicitly incorporate policy evolution dynamics and strategic adaptation as fundamental features rather than technical obstacles. Unlike standard MARL approaches that treat non-stationarity as a problem to overcome, our framework models how sophisticated agents reason about and deliberately influence others' learning trajectories \citep{kim2022influencing}, similar to the strategic teaching phenomena identified in economic experimentation literature \citep{yamamoto2019stochastic}.

Second, we incorporate partial observability as an intrinsic characteristic of social learning environments. By modeling belief states and observation functions, our approach captures the information asymmetries and strategic uncertainty that economic models identify as crucial determinants of learning outcomes \citep{heidhues2015strategic, rosenberg2009informational}.

Third, we account for long-horizon strategic planning where agents optimize not just immediate rewards but also their influence on the future learning dynamics of other agents. This aligns with economic perspectives on forward-looking behavior while remaining computationally tractable through reinforcement learning techniques.

The resulting framework enables more realistic modeling of social learning phenomena that resist analysis under either purely economic or purely computational approaches. It combines economic insights on strategic sophistication with MARL's ability to simulate complex adaptive systems, yielding both theoretical insights and practical algorithms for understanding multi-agent learning in partially observable environments.

\chapter{Theoretical Framework}
